---
title: "Class 5: Data viz with gplot2"
author: "Giselle A17995734"
format: pdf
---

#backwards

There are many GRAPHICS SYSTEMS IN R FOR MAKING PLOTS AND FIGURES. These include so-called *"base R " graphics* like the `plot()` fucntion and add on packages like **ggplot2**

let compare how we make a simple figure with these two systems:

We can use the built in cars dataset:

```{r}
plot(cars)
```

Before I can use ggplot 2 I need to install it on my computer. To do this we can use the function The main function in the **ggplot 2** package is called ggplot()

> **N.B.** We never `install.packages()` in our own quarto doc (We run it once only in our R cpnsole) as it would reinstall the package everytime we render our quarto report

once installed we need re load up the package

# Our first ggplot
# we need  data + aes +geoms
```{r}
ggplot(data=cars) +
aes(x=speed, y=dist) +
  geom_point()
```

```{r}
p <- ggplot(data=cars) +
    aes(x=speed, y=dist) +
    geom_point()
```

# Add a trend line

```{r}
p + geom_smooth()
p+ geom_smooth(method="lm")
```

# Read in our drug expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

#Lets make a first data plot attempt
#Add some color

```{r}
ggplot(data=genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
```

>Q. How Many genes ae in this dataset

>Q. How many 'up' regulated genes

```{r}
table( genes$State )
```

>Q. What fraction of a total genes is up regulated?
2.44




Make a plot for 1977 and 2007 (not all the years in the data set)

> First use the **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007

> Q. Make a histogram of lifeExp colored by continent (try suing `fill=continent` or `col=continent`) Q. Make a histogram of lifeExp faceted by continetnt





